---
title: "thesis"
author: "Milou Kleijn"
date: "`13/08/26`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Packages 
```{r}
library(readr)
library(tidyverse)
library(pROC)
library(randomForest)
library(gbm3)
library(writexl)
library(knitr)
library(extrafont)
library(effectsize)
```

# Dataset 
```{r}
ESS5 <- read_csv("ESS5_cleaned_allcountries_675.csv")
```

# Preperation
```{r}
# Filter Germany
ESS5.DE <- ESS5 %>% 
  filter(cntry == 'DE')

# Filter out minors under 18 and missings
ESS5.DE.filt <- ESS5.DE %>%
  filter(agea >= 18) %>% 
  filter(agea != 999999) %>% 
  filter(emplrel != 666666)

# Create binary dependent variable
ESS5.DE.STEM <- ESS5.DE.filt %>%
  mutate(stem = ifelse(
    (iscoco >= 2110 & iscoco <= 2114) |     # physicists, chemists etc.
      (iscoco >= 2120 & iscoco <= 2122) |   # mathematicians, statisticians
      (iscoco >= 2130 & iscoco <= 2139) |   # computing professionals
      (iscoco >= 2140 & iscoco <= 2146) |   # architects, engineers
      (iscoco >= 3110 & iscoco <= 3116) |   # physical/engineering technicians
      (iscoco >= 3121 & iscoco <= 3123),    # computer assistants/operators
    1, 0))


```

# Variable Selection (Decision 0)
```{r}
# Select only the variables I need
ESS5.DE.final <- ESS5.DE.STEM %>%
  select(idno, agea, emplrel, rsnlvem, yrspdwka, uemp3m, impdiff, atncrse, 
         netuse, sclmeet, sclact, health, hlthhmp, actvgrs, 
         edulvlb, ipeqopt, dscrgrp, dscrrce, dscretn, 
         wmcpwrk, mnrgtjb, dscrgnd, dscrage, imprich, ipfrule, 
         jbspv, stem)

```

# NA's
```{r}
# Check and remove missings (999999)
colSums(sapply(ESS5.DE.final, function(x) x == 999999), na.rm = TRUE)
ESS5.DE.final <- ESS5.DE.final[!rowSums(ESS5.DE.final == 999999) > 0, ]
ESS5.DE.final <- ESS5.DE.final %>%
  filter(impdiff != 8) %>%
  filter(ipeqopt != 8) %>% 
  filter(imprich != 8) %>% 
  filter(ipfrule != 8)

nrow(ESS5.DE.final)
```

## Recoding variables 
```{r}
# Recode binary variables to 0/1 instead of 1/2
ESS5.DE.final.rec <- ESS5.DE.final %>% 
  mutate(
    uemp3m = ifelse(uemp3m == 1, 1, 0), #If the value is 1, keep it as 1, 
    atncrse = ifelse(atncrse == 1, 1, 0), #anything else, make it 0 
    jbspv = ifelse(jbspv == 1, 1, 0), 
    dscrgrp = ifelse(dscrgrp == 1, 1, 0)
    ) %>% 
# Recode education variable into smaller groups based on ISCED 2011 categorization
  mutate(edulvlb = case_when(
    edulvlb == 5555 ~ "Other",
    edulvlb < 100 ~ "ISCED 0 - Less than primary",
    edulvlb < 200 ~ "ISCED 1 - Primary",
    edulvlb < 300 ~ "ISCED 2 - Lower secondary",
    edulvlb < 400 ~ "ISCED 3 - Upper secondary",
    edulvlb < 500 ~ "ISCED 4 - Post-secondary non-tertiary",
    edulvlb < 600 ~ "ISCED 5 - Short-cycle tertiary",
    edulvlb < 700 ~ "ISCED 6 - Bachelor or equivalent",
    edulvlb < 800 ~ "ISCED 7 - Master or equivalent",
    edulvlb < 900 ~ "ISCED 8 - Doctoral or equivalent"
  )) %>% 
   filter(edulvlb != "Other") %>% #Not a category in ISCED
  # Recode binary and nominal variables into factors
  mutate(across(c(emplrel, rsnlvem, uemp3m, atncrse, dscrgrp, dscrrce, dscretn,
                  dscrgnd, dscrage, hlthhmp, jbspv, stem, edulvlb), as.factor)
  ) 
```

# Final sample
```{r}
# Final sample size
nrow(ESS5.DE.final.rec)
table(ESS5.DE.final.rec$stem)
prop.table(table(ESS5.DE.final.rec$stem))
```

## Descriptives 

```{r}
mean(ESS5.DE.final.rec$agea)
sd(ESS5.DE.final.rec$agea)

as.data.frame(table(ESS5.DE.final.rec$edulvlb))
as.data.frame(table(ESS5.DE.final.rec$emplrel))
```

## Training set + Test set
```{r}
set.seed(1512)

train_index <- sample(1:nrow(ESS5.DE.final.rec), 0.8 * nrow(ESS5.DE.final.rec))

train_df <- ESS5.DE.final.rec[train_index, ]
test_df <- ESS5.DE.final.rec[-train_index, ]

# Check STEM distribution in both sets
prop.table(table(train_df$stem))
prop.table(table(test_df$stem))
```

### Gender for descriptive analysis (so not added as predictor)
```{r}
gender_lookup <- ESS5.DE.STEM %>% select(idno, gndr)

test_df_with_gender <- test_df %>%
  left_join(gender_lookup, by = "idno")

prop.table(table(test_df_with_gender$gndr))

```

## Bootstrap oversampling
```{r}
set.seed(1512)

stem_cases <- train_df[train_df$stem == 1, ]
non_stem_cases <- train_df[train_df$stem == 0, ]

stem_oversampled <- stem_cases[sample(nrow(stem_cases), 
                                      size = nrow(non_stem_cases), 
                                      replace = TRUE), ]

train_df_balanced <- rbind(non_stem_cases, stem_oversampled)
prop.table(table(train_df_balanced$stem))
```

# Age operationalization (Decision 1)
```{r}

operationalize_age <- function(df, type, cutoff = NULL) {
  if (type == "continuous") {
    return(df)  # agea already numeric, no change needed
  } else if (type == "binary") {
    df <- df %>%
      mutate(agea = ifelse(agea >= cutoff, 1, 0))
    return(df)
  }
}
```

# Algorithms (Decision 2)

### Algorithm 1: Multinomial logisitc regression
```{r}

fit_glm <- function(train_df, test_df) { 

  #Model
  glm_model <- glm(stem ~ . - idno, data = train_df, family = binomial)
  
  #Predictions
  pred_prob_train <- predict(glm_model, newdata = train_df, type = "response")
  pred_prob <- predict(glm_model, newdata = test_df, type = "response")
  
  y_train <- as.numeric(as.character(train_df$stem))
  y_test <- as.numeric(as.character(test_df$stem))
  #AUC
  auc_train <- roc(y_train, pred_prob_train) %>% auc()
  auc_test <- roc(y_test, pred_prob) %>% auc()
  
  #Output
  evaluation <- tibble(auc_train = auc_train, 
                       auc_test = auc_test)
  
  results <- list(evaluation = evaluation,
                  predictions_train = pred_prob_train, 
                  predictions_test = pred_prob,
                  model = glm_model
                   )

  return(results)
  }
```

### Algorithm 2: Bagging
```{r}

fit_bagging <- function(train_df, test_df) { 
  set.seed(1512)
  
  p <- ncol(train_df) - 1
  #Default no. trees = 500
  bag_model <- randomForest(stem ~ . - idno, data = train_df,
                            mtry = p,
                            ntree = 500, importance = TRUE)
  #Out-of-bag error estimation
  OOB_err_est <- bag_model$err.rate[bag_model$ntree, "OOB"]
  
  #Predictions
  pred_model <- predict(bag_model, newdata = test_df)
  pred_prob_train <- predict(bag_model, newdata = train_df, type = "prob") [,2]
  pred_prob <- predict(bag_model, newdata = test_df, type = "prob")[,2]
  
  y_train <- as.numeric(as.character(train_df$stem))
  y_test <- as.numeric(as.character(test_df$stem))
  
  #AUC
  auc_train <- roc(y_train, pred_prob_train) %>% auc()
  auc_test <- roc(y_test, pred_prob) %>% auc()
  
  #Output
  evaluation <- tibble(OOB_error = OOB_err_est,
                       auc_train = auc_train,
                       auc_test = auc_test)
  
  results <- list(evaluation = evaluation,
                  predictions_train = pred_prob_train, 
                  predictions_test = pred_prob,
                  model = bag_model
                  )
  
  return(results)
  }
```

### Algorithm 3: Boosting
```{r}

fit_boosting <- function(train_df, test_df) { 
  set.seed(1512)
  
  train_df <- train_df %>% mutate(stem = as.numeric(as.character(stem)))
  test_df <- test_df %>% mutate(stem = as.numeric(as.character(stem)))
  
  #Model 
  boost_model <- gbm(stem ~. - idno, data = train_df, 
                    distribution = "bernoulli", 
                    n.trees = 5000, 
                    interaction.depth = 4)
  
  #Predictions
  pred_prob_train <- predict(boost_model, newdata = train_df, n.trees = 5000, 
                             type = "response")
  pred_prob <- predict(boost_model, newdata = test_df, n.trees = 5000, 
                       type = "response")
  
  y_train <- as.numeric(as.character(train_df$stem))
  y_test <- as.numeric(as.character(test_df$stem))
  
  #AUC
  auc_train <- roc(y_train, pred_prob_train) %>% auc()
  auc_test <- roc(y_test, pred_prob) %>% auc()
  
  #Output
  evaluation <- tibble(auc_train = auc_train, 
                       auc_test = auc_test)
  
  results <- list(evaluation = evaluation,
                  predictions_train = pred_prob_train, 
                  predictions_test = pred_prob,
                  model = boost_model)
  
  return(results)
  }
```

# Threshols (Decision 3)

```{r}

apply_threshold <- function(pred_prob, threshold) {
  as.factor(ifelse(pred_prob >= threshold, 1, 0))

  }
```

# Combined function for all decisions

```{r}

run_models <- function(train_df, test_df, age_type, cutoff = NULL) {
  
  #Apply age operationalization
  train <- operationalize_age(train_df, age_type, cutoff)
  test <- operationalize_age(test_df, age_type, cutoff)
  
  #Run all three algorithms
  glm_results <- fit_glm(train, test)
  bag_results <- fit_bagging(train, test)
  boost_results <- fit_boosting(train, test)
  
  #Apply all three thresholds to test predictions
  thresholds <- c(0.3, 0.5, 0.7)
  
  classifications <- lapply(thresholds, function(thresh) {
    list(
      threshold = thresh,
      glm = apply_threshold(glm_results$predictions_test, thresh),
      bagging = apply_threshold(bag_results$predictions_test, thresh),
      boosting = apply_threshold(boost_results$predictions_test, thresh)
    )
  })
  
  return(list(
    glm = glm_results,
    bagging = bag_results,
    boosting = boost_results,
    classifications = classifications
  ))
}
```

# Results per age operationalization (50 used for the rest of the analysis)

```{r}
results_continuous <- run_models(train_df_balanced, test_df, "continuous")
results_age40 <- run_models(train_df_balanced, test_df, "binary", cutoff = 40)
results_age45 <- run_models(train_df_balanced, test_df, "binary", cutoff = 45)
results_age50 <- run_models(train_df_balanced, test_df, "binary", cutoff = 50)
results_age55 <- run_models(train_df_balanced, test_df, "binary", cutoff = 55)
```

# Models

```{r}
# Baseline
bin50_glm_05 <- apply_threshold(results_age50$glm$predictions_test, 0.5)
# Continuous age
cont_glm_03 <- apply_threshold(results_continuous$glm$predictions_test, 0.3)
cont_glm_05 <- apply_threshold(results_continuous$glm$predictions_test, 0.5)
cont_glm_07 <- apply_threshold(results_continuous$glm$predictions_test, 0.7)

cont_bag_03 <- apply_threshold(results_continuous$bagging$predictions_test, 0.3)
cont_bag_05 <- apply_threshold(results_continuous$bagging$predictions_test, 0.5)
cont_bag_07 <- apply_threshold(results_continuous$bagging$predictions_test, 0.7)

cont_boost_03 <- apply_threshold(results_continuous$boosting$predictions_test, 0.3)
cont_boost_05 <- apply_threshold(results_continuous$boosting$predictions_test, 0.5)
cont_boost_07 <- apply_threshold(results_continuous$boosting$predictions_test, 0.7)

# Binary age at 50
bin50_glm_03 <- apply_threshold(results_age50$glm$predictions_test, 0.3)
bin50_glm_07 <- apply_threshold(results_age50$glm$predictions_test, 0.7)

bin50_bag_03 <- apply_threshold(results_age50$bagging$predictions_test, 0.3)
bin50_bag_05 <- apply_threshold(results_age50$bagging$predictions_test, 0.5)
bin50_bag_07 <- apply_threshold(results_age50$bagging$predictions_test, 0.7)

bin50_boost_03 <- apply_threshold(results_age50$boosting$predictions_test, 0.3)
bin50_boost_05 <- apply_threshold(results_age50$boosting$predictions_test, 0.5)
bin50_boost_07 <- apply_threshold(results_age50$boosting$predictions_test, 0.7)

```

## Evaluate models function

```{r}

evaluate_model <- function(model_classifications, test_df, pred_prob) {
  
  true_labels <- as.factor(as.numeric(as.character(test_df$stem)))
  cm <- confusionMatrix(model_classifications, true_labels, positive = "1")
  auc <- roc(as.numeric(as.character(test_df$stem)), pred_prob) %>% auc()
  
  return(tibble(
    auc = as.numeric(auc),
    sensitivity = cm$byClass["Sensitivity"],
    specificity = cm$byClass["Specificity"]
  ))
}
```

# Evaluate models

```{r}
# Baseline: Binary age 50 - GLM - threshold 0.5
eval_bin50_glm_05 <- evaluate_model(bin50_glm_05, test_df,
                                    results_age50$glm$predictions_test)

# Continuous age - GLM
eval_cont_glm_03 <- evaluate_model(cont_glm_03, test_df, 
                                   results_continuous$glm$predictions_test)
eval_cont_glm_05 <- evaluate_model(cont_glm_05, test_df, 
                                   results_continuous$glm$predictions_test)
eval_cont_glm_07 <- evaluate_model(cont_glm_07, test_df, 
                                   results_continuous$glm$predictions_test)

# Continuous age - Bagging
eval_cont_bag_03 <- evaluate_model(cont_bag_03, test_df, 
                                   results_continuous$bagging$predictions_test)
eval_cont_bag_05 <- evaluate_model(cont_bag_05, test_df, 
                                   results_continuous$bagging$predictions_test)
eval_cont_bag_07 <- evaluate_model(cont_bag_07, test_df, 
                                   results_continuous$bagging$predictions_test)

# Continuous age - Boosting
eval_cont_boost_03 <- evaluate_model(cont_boost_03, test_df, 
                                     results_continuous$boosting$predictions_test)
eval_cont_boost_05 <- evaluate_model(cont_boost_05, test_df, 
                                     results_continuous$boosting$predictions_test)
eval_cont_boost_07 <- evaluate_model(cont_boost_07, test_df, 
                                     results_continuous$boosting$predictions_test)

# Binary age 50 - GLM
eval_bin50_glm_03 <- evaluate_model(bin50_glm_03, test_df, 
                                    results_age50$glm$predictions_test)
eval_bin50_glm_07 <- evaluate_model(bin50_glm_07, test_df, 
                                    results_age50$glm$predictions_test)

# Binary age 50 - Bagging
eval_bin50_bag_03 <- evaluate_model(bin50_bag_03, test_df, 
                                    results_age50$bagging$predictions_test)
eval_bin50_bag_05 <- evaluate_model(bin50_bag_05, test_df, 
                                    results_age50$bagging$predictions_test)
eval_bin50_bag_07 <- evaluate_model(bin50_bag_07, test_df, 
                                    results_age50$bagging$predictions_test)

# Binary age 50 - Boosting
eval_bin50_boost_03 <- evaluate_model(bin50_boost_03, test_df, 
                                      results_age50$boosting$predictions_test)
eval_bin50_boost_05 <- evaluate_model(bin50_boost_05, test_df, 
                                      results_age50$boosting$predictions_test)
eval_bin50_boost_07 <- evaluate_model(bin50_boost_07, test_df, 
                                      results_age50$boosting$predictions_test)

```

# Contradiction points

```{r}

find_contradictions <- function(p1, p2, T) {
  P <- data.frame(p1, p2)
  
  # Find cases where classification differs
  indices <- which((p1 < T & p2 > T) | (p1 > T & p2 < T))
  P_inbetween <- P[indices, ]
  P_n <- P_inbetween %>% nrow()
  P_per <- round(100 * P_n / length(p1), 1)
  
  found <- c("all" = length(p1), 
             "n_contradictions" = P_n, 
             "perc_contradictions" = P_per)
  
  outcome <- list(found, P_inbetween, indices)
  return(outcome)
}

```

### Contradiction points vs baseline

```{r}
# Baseline probabilities
baseline_prob <- results_age50$glm$predictions_test

# Continuous age - GLM
cp_cont_glm_03 <- find_contradictions(baseline_prob, results_continuous$glm$predictions_test, 0.3)
cp_cont_glm_05 <- find_contradictions(baseline_prob, results_continuous$glm$predictions_test, 0.5)
cp_cont_glm_07 <- find_contradictions(baseline_prob, results_continuous$glm$predictions_test, 0.7)

# Continuous age - Bagging
cp_cont_bag_03 <- find_contradictions(baseline_prob, results_continuous$bagging$predictions_test, 0.3)
cp_cont_bag_05 <- find_contradictions(baseline_prob, results_continuous$bagging$predictions_test, 0.5)
cp_cont_bag_07 <- find_contradictions(baseline_prob, results_continuous$bagging$predictions_test, 0.7)

# Continuous age - Boosting
cp_cont_boost_03 <- find_contradictions(baseline_prob, results_continuous$boosting$predictions_test, 0.3)
cp_cont_boost_05 <- find_contradictions(baseline_prob, results_continuous$boosting$predictions_test, 0.5)
cp_cont_boost_07 <- find_contradictions(baseline_prob, results_continuous$boosting$predictions_test, 0.7)

# Binary age 50 - GLM
cp_bin50_glm_03 <- find_contradictions(baseline_prob, results_age50$glm$predictions_test, 0.3)
cp_bin50_glm_07 <- find_contradictions(baseline_prob, results_age50$glm$predictions_test, 0.7)

# Binary age 50 - Bagging
cp_bin50_bag_03 <- find_contradictions(baseline_prob, results_age50$bagging$predictions_test, 0.3)
cp_bin50_bag_05 <- find_contradictions(baseline_prob, results_age50$bagging$predictions_test, 0.5)
cp_bin50_bag_07 <- find_contradictions(baseline_prob, results_age50$bagging$predictions_test, 0.7)

# Binary age 50 - Boosting
cp_bin50_boost_03 <- find_contradictions(baseline_prob, results_age50$boosting$predictions_test, 0.3)
cp_bin50_boost_05 <- find_contradictions(baseline_prob, results_age50$boosting$predictions_test, 0.5)
cp_bin50_boost_07 <- find_contradictions(baseline_prob, results_age50$boosting$predictions_test, 0.7)

```

# Combined table for evaluation

```{r}

results_table <- rbind(
  # Baseline
  tibble(age = "binary_50", algorithm = "GLM", threshold = 0.5,
         auc = eval_bin50_glm_05$auc,
         sensitivity = eval_bin50_glm_05$sensitivity,
         specificity = eval_bin50_glm_05$specificity,
         n_contradictions = 0, perc_contradictions = 0),
  
  # Continuous age - GLM
  tibble(age = "continuous", algorithm = "GLM", threshold = 0.3,
         auc = eval_cont_glm_03$auc,
         sensitivity = eval_cont_glm_03$sensitivity,
         specificity = eval_cont_glm_03$specificity,
         n_contradictions = cp_cont_glm_03[[1]]["n_contradictions"],
         perc_contradictions = cp_cont_glm_03[[1]]["perc_contradictions"]),
  
  tibble(age = "continuous", algorithm = "GLM", threshold = 0.5,
         auc = eval_cont_glm_05$auc,
         sensitivity = eval_cont_glm_05$sensitivity,
         specificity = eval_cont_glm_05$specificity,
         n_contradictions = cp_cont_glm_05[[1]]["n_contradictions"],
         perc_contradictions = cp_cont_glm_05[[1]]["perc_contradictions"]),
  
  tibble(age = "continuous", algorithm = "GLM", threshold = 0.7,
         auc = eval_cont_glm_07$auc,
         sensitivity = eval_cont_glm_07$sensitivity,
         specificity = eval_cont_glm_07$specificity,
         n_contradictions = cp_cont_glm_07[[1]]["n_contradictions"],
         perc_contradictions = cp_cont_glm_07[[1]]["perc_contradictions"]),
  
  # Continuous age - Bagging
  tibble(age = "continuous", algorithm = "Bagging", threshold = 0.3,
         auc = eval_cont_bag_03$auc,
         sensitivity = eval_cont_bag_03$sensitivity,
         specificity = eval_cont_bag_03$specificity,
         n_contradictions = cp_cont_bag_03[[1]]["n_contradictions"],
         perc_contradictions = cp_cont_bag_03[[1]]["perc_contradictions"]),
  
  tibble(age = "continuous", algorithm = "Bagging", threshold = 0.5,
         auc = eval_cont_bag_05$auc,
         sensitivity = eval_cont_bag_05$sensitivity,
         specificity = eval_cont_bag_05$specificity,
         n_contradictions = cp_cont_bag_05[[1]]["n_contradictions"],
         perc_contradictions = cp_cont_bag_05[[1]]["perc_contradictions"]),
  
  tibble(age = "continuous", algorithm = "Bagging", threshold = 0.7,
         auc = eval_cont_bag_07$auc,
         sensitivity = eval_cont_bag_07$sensitivity,
         specificity = eval_cont_bag_07$specificity,
         n_contradictions = cp_cont_bag_07[[1]]["n_contradictions"],
         perc_contradictions = cp_cont_bag_07[[1]]["perc_contradictions"]),
  
  # Continuous age - Boosting
  tibble(age = "continuous", algorithm = "Boosting", threshold = 0.3,
         auc = eval_cont_boost_03$auc,
         sensitivity = eval_cont_boost_03$sensitivity,
         specificity = eval_cont_boost_03$specificity,
         n_contradictions = cp_cont_boost_03[[1]]["n_contradictions"],
         perc_contradictions = cp_cont_boost_03[[1]]["perc_contradictions"]),
  
  tibble(age = "continuous", algorithm = "Boosting", threshold = 0.5,
         auc = eval_cont_boost_05$auc,
         sensitivity = eval_cont_boost_05$sensitivity,
         specificity = eval_cont_boost_05$specificity,
         n_contradictions = cp_cont_boost_05[[1]]["n_contradictions"],
         perc_contradictions = cp_cont_boost_05[[1]]["perc_contradictions"]),
  
  tibble(age = "continuous", algorithm = "Boosting", threshold = 0.7,
         auc = eval_cont_boost_07$auc,
         sensitivity = eval_cont_boost_07$sensitivity,
         specificity = eval_cont_boost_07$specificity,
         n_contradictions = cp_cont_boost_07[[1]]["n_contradictions"],
         perc_contradictions = cp_cont_boost_07[[1]]["perc_contradictions"]),
  
  # Binary age 50 - GLM
  tibble(age = "binary_50", algorithm = "GLM", threshold = 0.3,
         auc = eval_bin50_glm_03$auc,
         sensitivity = eval_bin50_glm_03$sensitivity,
         specificity = eval_bin50_glm_03$specificity,
         n_contradictions = cp_bin50_glm_03[[1]]["n_contradictions"],
         perc_contradictions = cp_bin50_glm_03[[1]]["perc_contradictions"]),
  
  tibble(age = "binary_50", algorithm = "GLM", threshold = 0.7,
         auc = eval_bin50_glm_07$auc,
         sensitivity = eval_bin50_glm_07$sensitivity,
         specificity = eval_bin50_glm_07$specificity,
         n_contradictions = cp_bin50_glm_07[[1]]["n_contradictions"],
         perc_contradictions = cp_bin50_glm_07[[1]]["perc_contradictions"]),
  
  # Binary age 50 - Bagging
  tibble(age = "binary_50", algorithm = "Bagging", threshold = 0.3,
         auc = eval_bin50_bag_03$auc,
         sensitivity = eval_bin50_bag_03$sensitivity,
         specificity = eval_bin50_bag_03$specificity,
         n_contradictions = cp_bin50_bag_03[[1]]["n_contradictions"],
         perc_contradictions = cp_bin50_bag_03[[1]]["perc_contradictions"]),
  
  tibble(age = "binary_50", algorithm = "Bagging", threshold = 0.5,
         auc = eval_bin50_bag_05$auc,
         sensitivity = eval_bin50_bag_05$sensitivity,
         specificity = eval_bin50_bag_05$specificity,
         n_contradictions = cp_bin50_bag_05[[1]]["n_contradictions"],
         perc_contradictions = cp_bin50_bag_05[[1]]["perc_contradictions"]),
  
  tibble(age = "binary_50", algorithm = "Bagging", threshold = 0.7,
         auc = eval_bin50_bag_07$auc,
         sensitivity = eval_bin50_bag_07$sensitivity,
         specificity = eval_bin50_bag_07$specificity,
         n_contradictions = cp_bin50_bag_07[[1]]["n_contradictions"],
         perc_contradictions = cp_bin50_bag_07[[1]]["perc_contradictions"]),
  
  # Binary age 50 - Boosting
  tibble(age = "binary_50", algorithm = "Boosting", threshold = 0.3,
         auc = eval_bin50_boost_03$auc,
         sensitivity = eval_bin50_boost_03$sensitivity,
         specificity = eval_bin50_boost_03$specificity,
         n_contradictions = cp_bin50_boost_03[[1]]["n_contradictions"],
         perc_contradictions = cp_bin50_boost_03[[1]]["perc_contradictions"]),
  
  tibble(age = "binary_50", algorithm = "Boosting", threshold = 0.5,
         auc = eval_bin50_boost_05$auc,
         sensitivity = eval_bin50_boost_05$sensitivity,
         specificity = eval_bin50_boost_05$specificity,
         n_contradictions = cp_bin50_boost_05[[1]]["n_contradictions"],
         perc_contradictions = cp_bin50_boost_05[[1]]["perc_contradictions"]),
  
  tibble(age = "binary_50", algorithm = "Boosting", threshold = 0.7,
         auc = eval_bin50_boost_07$auc,
         sensitivity = eval_bin50_boost_07$sensitivity,
         specificity = eval_bin50_boost_07$specificity,
         n_contradictions = cp_bin50_boost_07[[1]]["n_contradictions"],
         perc_contradictions = cp_bin50_boost_07[[1]]["perc_contradictions"])
)

results_table
write_xlsx(results_table, "results_table.xlsx")
```

## Age characteristics

```{r}
age_analysis <- function(cp_object, test_df_with_gender) {
  
  indices <- cp_object[[3]]
  
  #Ages of people whose classification changes 
  ages_change <- test_df_with_gender$agea[indices]
  ages_no_change <- test_df_with_gender$agea[-indices]
  
  #Ages of people whose classification does not change
  gender_change <- test_df_with_gender$gndr[indices]
  gender_no_change <- test_df_with_gender$gndr[-indices]
  
  return(tibble(
    n_change = length(ages_change),
    mean_age_change = round(mean(ages_change), 1),
    mean_age_no_change = round(mean(ages_no_change), 1),
    age_difference = round(mean(ages_change) - mean(ages_no_change), 1),
    pct_female_change = round(mean(gender_change == 2) * 100, 1)
  ))
}
```

```{r}
age_cont_glm_03 <- age_analysis(cp_cont_glm_03, test_df_with_gender)
age_cont_glm_05 <- age_analysis(cp_cont_glm_05, test_df_with_gender)
age_cont_glm_07 <- age_analysis(cp_cont_glm_07, test_df_with_gender)

age_cont_bag_03 <- age_analysis(cp_cont_bag_03, test_df_with_gender)
age_cont_bag_05 <- age_analysis(cp_cont_bag_05, test_df_with_gender)
age_cont_bag_07 <- age_analysis(cp_cont_bag_07, test_df_with_gender)

age_cont_boost_03 <- age_analysis(cp_cont_boost_03, test_df_with_gender)
age_cont_boost_05 <- age_analysis(cp_cont_boost_05, test_df_with_gender)
age_cont_boost_07 <- age_analysis(cp_cont_boost_07, test_df_with_gender)

age_bin50_glm_03 <- age_analysis(cp_bin50_glm_03, test_df_with_gender)
age_bin50_glm_07 <- age_analysis(cp_bin50_glm_07, test_df_with_gender)

age_bin50_bag_03 <- age_analysis(cp_bin50_bag_03, test_df_with_gender)
age_bin50_bag_05 <- age_analysis(cp_bin50_bag_05, test_df_with_gender)
age_bin50_bag_07 <- age_analysis(cp_bin50_bag_07, test_df_with_gender)

age_bin50_boost_03 <- age_analysis(cp_bin50_boost_03, test_df_with_gender)
age_bin50_boost_05 <- age_analysis(cp_bin50_boost_05, test_df_with_gender)
age_bin50_boost_07 <- age_analysis(cp_bin50_boost_07, test_df_with_gender)

```

```{r}
age_table <- rbind(
  #Baseline
  tibble(age = "binary_50", algorithm = "GLM", threshold = 0.5,
         n_change = 0, mean_age_change = NA, mean_age_no_change = NA, age_difference = NA,
         pct_female_change = NA),
  
  #Continuous age - GLM
  tibble(age = "continuous", algorithm = "GLM", threshold = 0.3,
         n_change = age_cont_glm_03$n_change,
         mean_age_change = age_cont_glm_03$mean_age_change,
         mean_age_no_change = age_cont_glm_03$mean_age_no_change,
         age_difference = age_cont_glm_03$age_difference,
         pct_female_change = age_cont_glm_03$pct_female_change),
  
  tibble(age = "continuous", algorithm = "GLM", threshold = 0.5,
         n_change = age_cont_glm_05$n_change,
         mean_age_change = age_cont_glm_05$mean_age_change,
         mean_age_no_change = age_cont_glm_05$mean_age_no_change,
         age_difference = age_cont_glm_05$age_difference,
         pct_female_change = age_cont_glm_05$pct_female_change),
  
  tibble(age = "continuous", algorithm = "GLM", threshold = 0.7,
         n_change = age_cont_glm_07$n_change,
         mean_age_change = age_cont_glm_07$mean_age_change,
         mean_age_no_change = age_cont_glm_07$mean_age_no_change,
         age_difference = age_cont_glm_07$age_difference,
         pct_female_change = age_cont_glm_07$pct_female_change),
  
  #Continuous age - Bagging
  tibble(age = "continuous", algorithm = "Bagging", threshold = 0.3,
         n_change = age_cont_bag_03$n_change,
         mean_age_change = age_cont_bag_03$mean_age_change,
         mean_age_no_change = age_cont_bag_03$mean_age_no_change,
         age_difference = age_cont_bag_03$age_difference,
         pct_female_change= age_cont_bag_03$pct_female_change),
  
  tibble(age = "continuous", algorithm = "Bagging", threshold = 0.5,
         n_change = age_cont_bag_05$n_change,
         mean_age_change = age_cont_bag_05$mean_age_change,
         mean_age_no_change = age_cont_bag_05$mean_age_no_change,
         age_difference = age_cont_bag_05$age_difference,
         pct_female_change = age_cont_bag_05$pct_female_change),
  
  tibble(age = "continuous", algorithm = "Bagging", threshold = 0.7,
         n_change = age_cont_bag_07$n_change,
         mean_age_change = age_cont_bag_07$mean_age_change,
         mean_age_no_change = age_cont_bag_07$mean_age_no_change,
         age_difference = age_cont_bag_07$age_difference,
         pct_female_change = age_cont_bag_07$pct_female_change),
  
  #Continuous age - Boosting
  tibble(age = "continuous", algorithm = "Boosting", threshold = 0.3,
         n_change = age_cont_boost_03$n_change,
         mean_age_change = age_cont_boost_03$mean_age_change,
         mean_age_no_change = age_cont_boost_03$mean_age_no_change,
         age_difference = age_cont_boost_03$age_difference,
         pct_female_change = age_cont_boost_03$pct_female_change),
  
  tibble(age = "continuous", algorithm = "Boosting", threshold = 0.5,
         n_change = age_cont_boost_05$n_change,
         mean_age_change = age_cont_boost_05$mean_age_change,
         mean_age_no_change = age_cont_boost_05$mean_age_no_change,
         age_difference = age_cont_boost_05$age_difference,
         pct_female_change = age_cont_boost_05$pct_female_change),
  
  tibble(age = "continuous", algorithm = "Boosting", threshold = 0.7,
         n_change = age_cont_boost_07$n_change,
         mean_age_change = age_cont_boost_07$mean_age_change,
         mean_age_no_change = age_cont_boost_07$mean_age_no_change,
         age_difference = age_cont_boost_07$age_difference,
         pct_female_change = age_cont_boost_07$pct_female_change),
  
  #Binary age 50 - GLM
  tibble(age = "binary_50", algorithm = "GLM", threshold = 0.3,
         n_change = age_bin50_glm_03$n_change,
         mean_age_change = age_bin50_glm_03$mean_age_change,
         mean_age_no_change = age_bin50_glm_03$mean_age_no_change,
         age_difference = age_bin50_glm_03$age_difference,
         pct_female_change= age_bin50_glm_03$pct_female_change),
  
  tibble(age = "binary_50", algorithm = "GLM", threshold = 0.7,
         n_change = age_bin50_glm_07$n_change,
         mean_age_change = age_bin50_glm_07$mean_age_change,
         mean_age_no_change = age_bin50_glm_07$mean_age_no_change,
         age_difference = age_bin50_glm_07$age_difference,
         pct_female_change = age_bin50_glm_07$pct_female_change),
  
  #Binary age 50 - Bagging
  tibble(age = "binary_50", algorithm = "Bagging", threshold = 0.3,
         n_change = age_bin50_bag_03$n_change,
         mean_age_change = age_bin50_bag_03$mean_age_change,
         mean_age_no_change = age_bin50_bag_03$mean_age_no_change,
         age_difference = age_bin50_bag_03$age_difference,
         pct_female_change = age_bin50_bag_03$pct_female_change),
  
  tibble(age = "binary_50", algorithm = "Bagging", threshold = 0.5,
         n_change = age_bin50_bag_05$n_change,
         mean_age_change = age_bin50_bag_05$mean_age_change,
         mean_age_no_change = age_bin50_bag_05$mean_age_no_change,
         age_difference = age_bin50_bag_05$age_difference,
         pct_female_change = age_bin50_bag_05$pct_female_change),
  
  tibble(age = "binary_50", algorithm = "Bagging", threshold = 0.7,
         n_change = age_bin50_bag_07$n_change,
         mean_age_change = age_bin50_bag_07$mean_age_change,
         mean_age_no_change = age_bin50_bag_07$mean_age_no_change,
         age_difference = age_bin50_bag_07$age_difference,
         pct_female_change= age_bin50_bag_07$pct_female_change),
  
  #Binary age 50 - Boosting
  tibble(age = "binary_50", algorithm = "Boosting", threshold = 0.3,
         n_change = age_bin50_boost_03$n_change,
         mean_age_change = age_bin50_boost_03$mean_age_change,
         mean_age_no_change = age_bin50_boost_03$mean_age_no_change,
         age_difference = age_bin50_boost_03$age_difference,
         pct_female_change = age_bin50_boost_03$pct_female_change),
  
  tibble(age = "binary_50", algorithm = "Boosting", threshold = 0.5,
         n_change = age_bin50_boost_05$n_change,
         mean_age_change = age_bin50_boost_05$mean_age_change,
         mean_age_no_change = age_bin50_boost_05$mean_age_no_change,
         age_difference = age_bin50_boost_05$age_difference,
         pct_female_change = age_bin50_boost_05$pct_female_change),
  
  tibble(age = "binary_50", algorithm = "Boosting", threshold = 0.7,
         n_change = age_bin50_boost_07$n_change,
         mean_age_change = age_bin50_boost_07$mean_age_change,
         mean_age_no_change = age_bin50_boost_07$mean_age_no_change,
         age_difference = age_bin50_boost_07$age_difference,
         pct_female_change = age_bin50_boost_07$pct_female_change)
  )

age_table
write_xlsx(age_table, "age_table.xlsx")
```

## Visualization (age)

```{r}

age_table %>%
  filter(n_change > 0) %>%
  mutate(config = paste(recode(age, "binary_50" = "Binary", "continuous" = "Continuous"),
                         algorithm, threshold, sep = " | "),
         config = reorder(config, mean_age_change)) %>%
  ggplot(aes(y = config)) +
  geom_segment(aes(x = mean_age_no_change, xend = mean_age_change, yend = config),
               color = "grey", linewidth = 0.8) +
  geom_point(aes(x = mean_age_no_change, color = "No contradiction points"), size = 3) +
  geom_point(aes(x = mean_age_change, color = "Contradiction points"), size = 3) +
  scale_color_manual(values = c("Contradiction points" = "#4a3aa7",
                                 "No contradiction points" = "#898781")) +
  labs(x = "Mean age (years)", y = NULL, color = NULL) +
  theme_minimal(base_size = 12, base_family = "Times New Roman") +
theme(
  panel.grid.minor = element_blank(),
  legend.position = "top",
  text = element_text(color = "black"),
  axis.text = element_text(color = "black"),
  axis.title = element_text(color = "black"),
  legend.text = element_text(color = "black"),
  legend.title = element_text(color = "black")
)
  theme(panel.grid.minor = element_blank(), legend.position = "top")

ggsave("mean_age_cp_vs_no_cp.png", width = 8, height = 7, dpi = 300)
```

# Correlations between age and numeric variables 

```{r}

cont_vars <- c("yrspdwka", "netuse", "sclmeet", "sclact", 
               "impdiff", "ipeqopt", "imprich", "ipfrule")

cont_results <- lapply(cont_vars, function(var_name) {
  test <- cor.test(ESS5.DE.final.rec$agea, 
                    ESS5.DE.final.rec[[var_name]], 
                    use = "pairwise.complete.obs")
  tibble(
    variable = var_name,
    type = "continuous",
    correlation = round(test$estimate, 3),
    p_value = round(test$p.value, 4)
  )
}) %>% bind_rows()

cont_results

```

```{r}

binary_vars <- c("uemp3m", "atncrse", "jbspv", "dscrgrp",
                  "dscrrce", "dscretn", "dscrgnd", "dscrage", "hlthhmp")

binary_results <- lapply(binary_vars, function(var_name) {
  var_numeric <- as.numeric(as.character(ESS5.DE.final.rec[[var_name]]))
  test <- cor.test(ESS5.DE.final.rec$agea, 
                    var_numeric, 
                    use = "pairwise.complete.obs")
  tibble(
    variable = var_name,
    type = "binary",
    correlation = round(test$estimate, 3),
    p_value = round(test$p.value, 4)
  )
}) %>% bind_rows()

binary_results

```

```{r}
comb <- bind_rows(cont_results, binary_results)
write_xlsx(comb, "correlation_matrix.xlsx")
```

# Proportion of variance in age explained by multi categorical variables

```{r}

multi_vars <- c("edulvlb", "emplrel", "rsnlvem")

multi_results <- lapply(multi_vars, function(var_name) {
  test <- aov(ESS5.DE.final.rec$agea ~ ESS5.DE.final.rec[[var_name]])
  summary_test <- summary(test)[[1]]
  eta_result <- eta_squared(test)
  
  tibble(
    variable = var_name,
    type = "multi-category",
    df1 = summary_test$Df[1],        # numerator df (between groups)
    df2 = summary_test$Df[2],        # denominator df (residual/within groups)
    f_statistic = round(summary_test$`F value`[1], 3),
    p_value = round(summary_test$`Pr(>F)`[1], 4),
    eta_squared = round(eta_result$Eta2[1], 3)
  )
}) %>% bind_rows()

multi_results
write_xlsx(multi_results, "multi_results.xlsx")

```

## Age coefficients 

```{r}
# For the continuous age operationalization
glm_model_cont <- results_continuous$glm$model
summary(glm_model_cont)

# For the binary (age 50 cutoff) operationalization
glm_model_bin50 <- results_age50$glm$model
summary(glm_model_bin50)
```

# Assumption checking - Log Linearity 
```{r}
#### Linearity

empirical_logit_by_value <- function(data, predictor, outcome = "stem") {
  df <- data %>%
    mutate(y = as.numeric(as.character(.data[[outcome]]))) %>%
    group_by(.data[[predictor]]) %>%
    summarise(
      n = n(),
      events = sum(y),
      emp_logit = log((events + 0.5) / (n - events + 0.5))
    )

  ggplot(df, aes(x = .data[[predictor]], y = emp_logit)) +
    geom_point(size = 3) +
    geom_line() +
    geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
    labs(title = paste("Empirical logit:", predictor),
         x = predictor, y = "Empirical log-odds of STEM classification") +
    theme_minimal(base_size = 12)
}

empirical_logit_plot <- function(data, predictor, outcome = "stem", n_bins = 10) {
  df <- data %>%
    mutate(y = as.numeric(as.character(.data[[outcome]])),
           bin = ntile(.data[[predictor]], n_bins)) %>%
    group_by(bin) %>%
    summarise(
      x_mean = mean(.data[[predictor]], na.rm = TRUE),
      n = n(),
      events = sum(y),
      emp_logit = log((events + 0.5) / (n - events + 0.5))
    )

  ggplot(df, aes(x = x_mean, y = emp_logit)) +
    geom_point(size = 3) +
    geom_line() +
    geom_smooth(method = "lm", se = FALSE, color = "red", linetype = "dashed") +
    labs(title = paste("Empirical logit:", predictor),
         x = predictor, y = "Empirical log-odds of STEM classification") +
    theme_minimal(base_size = 12)
}

# Graph per predictor 
p1 <- empirical_logit_by_value(train_df_balanced, "sclmeet")
ggsave("logit_sclmeet.png", plot = p1, width = 6, height = 4, dpi = 300)

p2 <- empirical_logit_by_value(train_df_balanced, "sclact")
ggsave("logit_sclact.png", plot = p2, width = 6, height = 4, dpi = 300)

p3 <- empirical_logit_by_value(train_df_balanced, "impdiff")
ggsave("logit_impdiff.png", plot = p3, width = 6, height = 4, dpi = 300)

p4 <- empirical_logit_by_value(train_df_balanced, "ipeqopt")
ggsave("logit_ipeqopt.png", plot = p4, width = 6, height = 4, dpi = 300)

p5 <- empirical_logit_by_value(train_df_balanced, "imprich")
ggsave("logit_imprich.png", plot = p5, width = 6, height = 4, dpi = 300)

p6 <- empirical_logit_by_value(train_df_balanced, "ipfrule")
ggsave("logit_ipfrule.png", plot = p6, width = 6, height = 4, dpi = 300)

p7 <- empirical_logit_by_value(train_df_balanced, "health")
ggsave("logit_health.png", plot = p7, width = 6, height = 4, dpi = 300)

p8 <- empirical_logit_by_value(train_df_balanced, "actvgrs")
ggsave("logit_actvgrs.png", plot = p8, width = 6, height = 4, dpi = 300)

p9 <- empirical_logit_by_value(train_df_balanced, "wmcpwrk")
ggsave("logit_wmcpwrk.png", plot = p9, width = 6, height = 4, dpi = 300)

p10 <- empirical_logit_by_value(train_df_balanced, "mnrgtjb")
ggsave("logit_mnrgtjb.png", plot = p10, width = 6, height = 4, dpi = 300)

p11 <- empirical_logit_by_value(train_df_balanced, "netuse")
ggsave("logit_netuse.png", plot = p11, width = 6, height = 4, dpi = 300) 

p12 <- empirical_logit_plot(train_df_balanced, "yrspdwka")
ggsave("logit_yrspdwka.png", plot = p12, width = 6, height = 4, dpi = 300)

p13 <- empirical_logit_plot(train_df_balanced, "agea")
ggsave("logit_agea.png", plot = p13, width = 6, height = 4, dpi = 300)

```
